---
title: "Parse BLAST Results"
author: "Philipp Ross"
date: "2018-01-10"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---

Here we want to parse the TBLASTX results such that we remove spurious alignments and determine the reciporical best hits. We will filter the alignments by a "standard" and "conservative" criteria.

```{r}
# load libraries
library(tidyverse)
library(magrittr)
```

## Filtering initial BLAST results

First we need to split up the query and subject IDs so we're matching just on Ensmbl gene IDs.

```{r,engine='bash',eval=F}
cat thirdset.txt | awk '{split($1,x,"|");split($2,y,"|");  print x[1],y[1],$3,$4,$5,$6,$7,$8,$9,$10}' > thirdset_processed.txt
```

Now we parse the BLAST results, filter them, and write those to an output file.

```{r}
# read in file
data <- readr::read_tsv("../data/thirdset_processed.txt",
                        comment = "#", 
                        col_names=c("query","subject","evalue","alnlength",
                                    "identity","identical","mismatches",
                                    "qframe","sframe","bframe")) %>%
  dplyr::filter(!(query == subject)) %>% # query does not equal subject
  dplyr::filter(!(qframe == sframe)) %>% # frames are not equal
  dplyr::filter(identity >= 80.0) %>% # minimum percentage identity
  dplyr::filter(evalue <= 1e-10) # minimum E-value

# get maximum alignment length
aln_lookup_table <- data %>%
  dplyr::group_by(query, subject) %>% 
  dplyr::summarise(max_aln_length = max(alnlength))

# filter by maximum alignment length
data <- dplyr::inner_join(data, aln_lookup_table, 
                          by = c("query","subject","alnlength"="max_aln_length"))

# get maximum percent identity
id_lookup_table <- data %>%
  dplyr::group_by(query, subject) %>% 
  dplyr::summarise(max_identity = max(identity))

# filter by maximum percent identity
data <- dplyr::inner_join(data, id_lookup_table, 
                          by = c("query","subject","identity"="max_identity"))
  
# write standard file
standard <- data %>%
  dplyr::filter(alnlength >= 50) # minimum sequence alignment length
  
readr::write_tsv(x = standard, path="../output/thirdset/standardset.tsv")

# write conservative file
conservative <- data %>%
  dplyr::filter(alnlength >= 100) # minimum sequence alignment length
  
readr::write_tsv(conservative, path="../output/thirdset/conservativeset.tsv")
```

## Finding reciprocal best hits for the conservative set

Now to find only the reciprocal best hits. We do this by making sure the best match for the query gene is also the best match for the subject gene. They each need to have the highest percent identity for each other and have complementary frame shifts. An aligment of query to subject with frame shifts of 1 and 3 respectively, should have reciprocal frame shifts of 3 and 1 or -3 and -1 or -1 and -3 or 1 and 3.

We do this using the not so pretty code below:

```{r}
# make an empty table placeholder
otable <- tibble::tibble(query=character(),
                           subject=character(),
                           evalue=double(),
                           alnlength=integer(),
                           identity=double(),
                           identical=integer(),
                           mismatches=integer(),
                           qframe=integer(),
                           sframe=integer())

# first make sure we're looping through IDs that are both in subject or query
interIDs <- dplyr::intersect(conservative$query, conservative$subject)

# make a dump list - anything that doesn't meet the criteria is added to this list
# if any future matches come up and either ID is in this list, don't add it
dump_list <- c()

# go through each ID in our intersected list
for (i in 1:length(interIDs)) {
  
  # loop through each query ID
  q <- interIDs[i]
  
  # if the ID is already in the dump list, go to the next one
  ifelse(q %in% dump_list, next(), TRUE)
  
  # create tables of every instance of this ID found in the query column
  # and in the subject column
  qtable <- dplyr::filter(conservative, query == q)
  stable <- dplyr::filter(conservative, subject == q)
  
  # take product of the two frames for comparison
  qqframe <- qtable$qframe
  qsframe <- qtable$sframe
  sqframe <- stable$qframe
  ssframe <- stable$sframe
  qtable$frameprod <- qqframe * qsframe
  stable$frameprod <- sqframe * ssframe
  
  # only keep the entries where the product is the same - 
  # we assume that this likely signifies a similar event for reciprocal
  # alignments
  tmp <- dplyr::inner_join(qtable, stable, by = "frameprod")
  
  # split back into two tables
  qtable <- dplyr::select(tmp, c(query=query.x,
                                 subject=subject.x,
                                 evalue=evalue.x,
                                 alnlength=alnlength.x,
                                 identity=identity.x,
                                 identical=identical.x,
                                 mismatches=mismatches.x,
                                 qframe=qframe.x,
                                 sframe=sframe.x))
  stable <- dplyr::select(tmp, c(query=query.y,
                                 subject=subject.y,
                                 evalue=evalue.y,
                                 alnlength=alnlength.y,
                                 identity=identity.y,
                                 identical=identical.y,
                                 mismatches=mismatches.y,
                                 qframe=qframe.y,
                                 sframe=sframe.y))
  
  # if this is empty, next iteration - this indicates the frameshifts are not
  # equivalent
  if (nrow(tmp)>0) {
         qmax_id <- max(qtable$identity)
         smax_id <- max(stable$identity)
  }
  else {
    next()
  }

  # only keep the entries where you have the maximum percent identity
  qtable <- dplyr::filter(qtable, identity == qmax_id)
  stable <- dplyr::filter(stable, identity == smax_id)
  
  # add entries to the dump list where the query aligns to a subject, but the
  # subject aligns better to another gene. These are very similar to one another
  # but are not reciprocal best hits to one another
  if (length(intersect(qtable$subject,stable$query))>0) {
    TRUE
  }
  else {
    dump_list <- c(dump_list, stable$query, qtable$subject)
    next()
  }
  
  # finally, merge rows
  otable <- dplyr::bind_rows(otable,qtable,stable)
}

# and only keep unique rows
otable <- unique(otable)

# sometimes duplicates of nearly reciprocal best hits remain
# lets remove the duplicate queries
dup_queries <- unique(otable[,1:2])[duplicated(unique(otable[,1:2])$query),]$query

# temporary place holder
dupqtable <- tibble::tibble(query=character(),
                           subject=character(),
                           evalue=double(),
                           alnlength=integer(),
                           identity=double(),
                           identical=integer(),
                           mismatches=integer(),
                           qframe=integer(),
                           sframe=integer())

for (i in 1:length(dup_queries)) {
  dq <- dup_queries[i]
  out <- dplyr::filter(otable, query == dq) %>% dplyr::arrange(desc(identity))
  dupqtable <- dplyr::bind_rows(dupqtable,out[2:nrow(out),])
}

# now lets remove the duplicate subjects
dup_subjects <- unique(otable[,1:2])[duplicated(unique(otable[,1:2])$subject),]$subject

# temporary placeholder
dupstable <- tibble::tibble(query=character(),
                           subject=character(),
                           evalue=double(),
                           alnlength=integer(),
                           identity=double(),
                           identical=integer(),
                           mismatches=integer(),
                           qframe=integer(),
                           sframe=integer())

for (i in 1:length(dup_subjects)) {
  ds <- dup_subjects[i]
  out <- dplyr::filter(otable, subject == ds) %>% dplyr::arrange(desc(identity))
  dupstable <- dplyr::bind_rows(dupstable,out[2:nrow(out),])
}

# remove those duplicates
otable <- dplyr::anti_join(otable,dupqtable)
otable <- dplyr::anti_join(otable,dupstable)
# remove things that were added to dump list, but due to ordering of queries
# and matches, they were not removed
otable <- dplyr::filter(otable, !(query %in% dump_list))
otable <- dplyr::filter(otable, !(subject %in% dump_list))
# the union and intersection of these lists should be equal
reciprocal_best_hits <- unique(otable[,1:2])
length(intersect(reciprocal_best_hits[[1]],reciprocal_best_hits[[2]]))
length(union(reciprocal_best_hits[[1]],reciprocal_best_hits[[2]]))
# these should contain the same elements
sum(intersect(reciprocal_best_hits[[1]],reciprocal_best_hits[[2]]) == union(reciprocal_best_hits[[1]],reciprocal_best_hits[[2]]))
```

```{r}
# write to an output file
readr::write_tsv(x=otable,path="../output/thirdset/conservative_otable.tsv")
readr::write_tsv(x=reciprocal_best_hits,path="../output/thirdset/conservative_reciprocal_best_hits.tsv")
```

How many frameshift events in our conservative set?

```{r}
length(reciprocal_best_hits[[1]])/2
```

How many genes in our conservative set?

```{r}
length(reciprocal_best_hits[[1]])
```

## Total events in standard set

Now we do something similar for the standard set except this time we aren't interested in keeping only best reciprocal hits, but all hits.

```{r}
# make an empty list to store results
standard_reciprocals <- c()
# loop through every row - this works because if the gene is not in the
# query column, then it obviously is not a reciprocal
for (i in 1:nrow(standard)) {
  # take query and subject of this row
  q <- standard$query[i]
  s <- standard$subject[i]
  # if the query is also in the subject column, continue
  if (q %in% standard$subject) {
    # find what the queries are for where this gene is a subject
    alt <- dplyr::filter(standard, subject==q) %$% query
    # if the subject was a query of the original query subject match, then
    # we keep it
    if (s %in% alt) {
      standard_reciprocals <- c(standard_reciprocals,q)
    } else {
      next()
    }
  } else {
    next()
  }
}
```

Now we should have our list of reciprocal best hits! For the standard set, we don't care about the reciporical best hits so we will just remove duplicate rows.

```{r}
standard_hits <- standard %>% 
  dplyr::filter(query %in% standard_reciprocals & subject %in% standard_reciprocals) %>%
  dplyr::select(query, subject) %>% 
  dplyr::distinct()
```

```{r}
# write to an output file
readr::write_tsv(x=standard_hits,path="../output/thirdset/standard_hits.tsv")
```

How many frameshift events in our standard set?

```{r}
# first remove all reciprocal best hits
anti <- dplyr::anti_join(standard_hits,reciprocal_best_hits)
# set count to zero and make an empty list
count <- 0
hits <- list()
# loop through table lacking reciprocal best hits making sure we
# count cases of A to B and B to A as one event, but
# cases of A to B and A to C as two events and A to B and B to C as two events
for (i in 1:nrow(standard_hits)) {
  row <- standard_hits[i,]
  if (is.null(hits[[row$query]])) {
    if (is.null(hits[[row$subject]])) {
      count <- count + 1
      hits[[row$query]] <- row$subject
      next()
    } else if (row$query %in% hits[[row$subject]]) {
      hits[[row$query]] <- row$subject
      next()
    } else {
      count <- count + 1
      hits[[row$query]] <- row$subject
      next()
    }
  } else if (row$subject %in% hits[[row$query]]) {
    next()
  } else {
    count <- count + 1
    hits[[row$query]] <- c(hits[[row$query]],row$subject)
    next()
  }
}
# add the count plus the number of reciprocal best hits divided in half to 
# get the total count of standard frameshifts
print(count)
```

Just to be sure, if we join the standard table with the conservative table, do we get everything that we have in the conservative table?

```{r}
length(reciprocal_best_hits[[1]]) == length(dplyr::inner_join(standard_hits,reciprocal_best_hits)[[1]])
```

How many genes in our standard set?

```{r}
length(unique(standard_hits$query,standard_hits$subject))
```
